An Integral Projection Modeling Approach to Understanding Demographic Effects of Multispecies Mutualisms

### 2024-01-21 Manuscript/ant_cactus_AmNat.tex
manuscript comments

\section*{Abstract}
\tom{}{I think this is too long for Am Nat requirements. Also, use ``we''.}
Mutualisms are widespread species interactions with diverse and dynamic consequences. 
They are considered more context dependent than other species interactions, meaning there are many different factors which change the outcomes of interactions between mutualists, including partner diversity. 
Partner diversity has become a central focus in the field of mutualisms, expanding previous work from primarily pairwise to multispecies mutualisms. 
It has been shown that pairwise studies are poor predictors of the effects of multispecies mutualistic interactions. 
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.

@@

\noindent Functions \tom{$P(x';\pmb{\tau^P})$ and $F(x';\pmb{\tau^F})$}{These should be $x$ not $x'$ because we modeled flowering and fertility in year t based on size in year t.} give the probability of flowering and the number of flowerbuds produced, respectively, by plants of size $x'$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit (\tom{$\kappa(a')$}{Why is this $a'$? I think it should be $a$. More generally, I think you need to explain your use of primes and how readers should interpret them at the start of model exposition.}) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that \tom{deviates are specific to plants in ant state $a$}{Why does the tau vector for viability have the $a$ subscript but the others do not?}. 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using Stan and Rstan (\tom{}{Cite Stan and RStan.}). 
Unless otherwise mentioned, all models use vague priors.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given the fixed effects of previous size $x$ and previous ant partner $a$ and random effects of plot $w$ and year $u$. 
We fit this model to size data at the end of the transition year $y^G$ using the location-scale parameterization of the student $t$ distribution, because in preliminary analyses we found that size transition data were more fat-tailed than a Gaussian distribution could accommodate. 
Specifically, the model was: 
$$y^G \sim Student T(\hat{nu},\hat{G},\hat{\sigma}) $$
$$\hat{G} = \beta_{0}^{G} + \beta_{1}^{G} \times x  + \beta_{2}^{G} \times a + \beta_{3}^{G} \times x \times a  + \beta_{4}^{G} \times x^2 + \beta_{5}^{G} \times a \times x^2 + u + w $$
$$\hat{\sigma} = \beta_{0}^{\sigma} + \beta_{1}^{\sigma} \times x $$
$$\hat{\nu} = \beta_{0}^{\nu} + \beta_{1}^{\nu} \times x $$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ with ant $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
Ants are included as predictors here because ant partners defend cholla from herbivores, preventing the loss of limbs and therefore shrinking.

\paragraph{Survival}
The survival model ($S(a,x;\pmb{\tau_{a}^{S}})$) estimates the probability of survival $y^S$ from year $t$ to year $t+1$, with fixed effects of the previous size of the cholla $x$ and ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^S \sim Bern(\hat{S})$$
$$logit(\hat{S} = \beta_{0}^{S}) + \beta_{1}^{S} \times x + \beta_{2}^{S} \times a + \beta_{3}^{S} \times x \times a + u + w$$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.
A Bernoulli distribution was chosen here because the only two possible outcomes are survival or death. 
Ants are included as predictors here because ant partners defend cholla from herbivores and predators, decreasing the likelihood of mortality due to either of these.

\paragraph{Reproduction}
The reproduction model ($P(x';\pmb{\tau^{P}})$) estimates the probability of reproducing $y^P$ in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^P \sim Bern(\hat{P})$$
$$logit(\hat{P}) = \beta_{0}^{P} + \beta_{1}^{P} \times x' + u + w$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Flowers}
The total flowers model ($F(x';\pmb{\tau^{F}})$) estimates the total flowers produced by a plant $y^F$ in year $t+1$, with fixed effects of size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a Negative Binomial distribution:
$$y^{F} \sim 0 Truncated Negative Binom(\hat{F},\hat{\phi})$$
$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u + w$$
$$log(\hat{\phi}) = \beta_{0}^{\phi}$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Viability}
The viability model ($V(a;\pmb{\tau^{V}_{a}})$) estimates the proportion of flowers produced by a plant which are viable (not aborted) $y^V$ in year $t+1$, with fixed effects of the ant partner of the cactus $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Binomial distribution:
$$y^{V} \sim Binom(\hat{V})$$
$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.
Ants are included as predictors here because they defend the cacti from seed predation which can lead to floral abortion.

\paragraph{Ant Transitions}
The ant transition rates model ($\epsilon(x,a,a';\pmb{\tau^{\epsilon}})$) estimates the probability of a cactus being visited by an ant partner $a'$ $y^{\epsilon}$, with fixed effects of the previous size of the cholla $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$, using a Multinomial distribution: 
$$y^{\epsilon} ~ Multinomial(\hat{\epsilon})$$
$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$.
Ant partners are included as predictors here because partners may choose to return to the same cholla repeatedly or choose new ones, therefore the previous partner may be a good indicator of the next partner.

\paragraph{Recruit Size Distribution}
The recruit size model ($n(x,a')$) estimates the size distribution of all recruits $y^{\eta}$ from a given year $t+1$, with no fixed or random effects, using a Normal distribution: 
$$y^{\eta} ~\sim N(\hat{\eta},\hat{\sigma})$$
$$\hat{\eta} = \beta_{0}^{\eta}$$
where $\hat{\sigma}$ is estimated with a non-informative prior.

\paragraph{Germination}
With germination data from \tom{Miller et al., 2007}{CITE}, we fit two Bayesian generalized linear models for the probability of germinating from a seed in the first year ($\gamma_1$) or the second year ($\gamma_2$) in year $t+1$, with no fixed or random effects, using a Binomial distribution:
$$y^{\gamma_1} \sim Binomial(\hat{\gamma_1})$$
$$y^{\gamma_2} \sim Binomial(\hat{\gamma_2})$$
$$logit(\hat{\gamma_1}) = \beta_{0}^{\gamma_1}$$
$$logit(\hat{\gamma_2}) = \beta_{0}^{\gamma_2}$$

\paragraph{Pre-Census Survival}
With data collected in a \tom{2005-2006}{FIX}********** recruit census, we fit a Bayesian generalized linear model for the probability of a seedling surviving to May ($\delta$) of year $t+1$ (accounting for missed mortality events), with fixed effects of the previous size $x$ and random effects of the transect $m$, using a Bernoulli distribution: 
$$y^{\delta} ~ Bern(\hat{\delta})$$
$$logit(\hat{\delta}) = \beta_{0}^{\delta} + m$$
 where $m \sim N(0, \sigma_{transect}^2)$ is the random effect of transect where the recruited individual was analyzed for survival.

\paragraph{Seeds Per Flower}
With data from \tom{Miller 2007}{CITE}********, we fit a Bayesian generalized linear model for the number of seeds produced by every flower on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$, using a Negative Binomial distribution:
$$y^{\kappa} \sim 0 Truncated Negative Binomial(\hat{\kappa},\hat{\phi})$$
$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
Ant partners are included as predictors here because they reduce floral abortion rates and therefore may lead to higher numbers of seeds. 
\tom{}{This paragraph needs to address how we assigned seed number to the ``other'' category.}

\paragraph{MCMC Simulations}
To obtain posterior estimates of the demographic parameters, we fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite} 
For each model, we obtained 3 chains of 10,000 iterations, each with randomly chosen initial conditions. 
The first 1,500 iterations were discarded as burn-in to eliminate transience associated with initial conditions. 
We did not thin the chains, thus all samples were retained. 
To assess the convergence of our models we assessed between and within chain convergence, the resulting figures are included in supplemental documents. 
To assess the overall model fit we carried out posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data.
Large differences in the two indicate a poor model fit and can be assessed visually (figures included in supplemental documents). 
All estimated parameters are available in the data. 
Data and code for all vital rate models is included in the supplemental information.

\subsection*{IPM Analysis}

Analyzing an IPM requires discretizing the composite IPM into a matrix to calculate the dominant eigenvalue. 
In a traditional IPM $x$ is discretized into $b$ bins, replacing the continuous kernel into a $b \times b$ matrix, in this case there is additional complexity in the form of transitions between ant partners. 
Each combination of previous ant state $a$ and next ant state $a'$ represents a unique set of plants in the popualtion and therefore must be discritized individually, leading to a matrix size of $m b \times m b$ where $m$ is the number of unique combinations of $a$ and $a'$ (how many possible ant transitions there are). 
In this model we have two additional discrete states (year one and year two seed banks) leading to a final matrix size of $m(b+2) \times m(b+2)$.
We used $b = 200$ bins.
We extend the integration limits $L$ and $U$ to avoid unintentional eviction \cite{Williams2012}.
Traditionally, in a deterministic IPM the asympotic population growth rate $\lambda$ is estimated as the dominant eigenvalue of the discrete kernel.
In a deterministic bayesian IPM, we create 1,000 discrete kernels with a unique set of parameters from our bayesian statistical models to estimate 1,000 $\lambda$ values as a distribution. 
In a deterministic version of our IPM, this process is repeated separately for every combination of ant partners: complete vacancy; \textit{C. opuntiae} and vacancy; other and vacancy; \textit{L. apiculatum}, \textit{C. opuntiae}, and vacancy, \textit{L. apiculatum}, other, and vacancy; \textit{C. opuntiae}, other, and vacancy; and all ant partners and vacancy.
In order to calculate stochastic $\lambda$ distributions for each ant scenario, every one of the 1,000 parameter-iteration associated $\lambda$s actually comes from the mean of 5,000 year-specific estimations each associated with a year random effech which was randomly selected from the 18 years of data.

We compare the distributions of $\lambda$ across each combination of ant partners to whether sampling effect or complementarity is at play in the system.
To compare the distributions of two hypothetical populations, A and B, we subtract the vector of 1,000 $\lambda_A$ estimations of one population from the 1,000 $\lambda_B$ estimations of the other ($\lambda_A - \lambda_B$).
If the average of these differences is positive, population A has a higher $\lambda$ than B, if the average is 0, they are equal, and if the average is negative, population B has a higher $\lambda$ than A.

To determine if sampling effect is at play in a system, we must first determine if there is a "best" partner, by determining which single ant association is correlated with the highest $\lambda$ estimation.
We can do this by comparing the distributions of each $\lambda$ and finding the one which is larger than all others.
Then we must show that the $\lambda$ of a population with all possible partners is equal to that of $\lambda$ of a population with only the "best" partner, by comparing the distributions as described earlier.
If it is significantly larger then complementarity is potentially at play (more on this in the next paragraph).
If it is significantly smaller, this indicates that rather than positive effects of partner diversity, there are actually important costs of partner diversity that dampen the population growth rate.

@@

To determine if portfolio effect is at play we must show that partner diversity buffers the population from environmental variation.
To do this we parameterized a new set of bayesian statistical models in which year random effects were not ant specific, meaning the effects of ant partners were not able to covary with year effects (the scenario in which all ant partners must respond to annual variation the same way and portfolio effect is not possible).
The new estimations of year random effects are $u \sim N(0,\sigma_{yr}^2)$ with year specific variance of $\sigma_{yr}$. 
With this verision of the IPM (now referred to as the stochastic null IPM) we followed the same approach to calculate distributions of stochastif $\lambda$s for each ant scenario.
We can estimate the effects of partner diversity on the focal mutualist by calculating the difference in $\lambda$ distributions between populations with no partners and populations with all partners present.
Then we can compare the difference in the effects of partner diversity on the focal mutualist calculated from the stochastic and stochastic null IPMs respectively.
If the stochastic difference is greater than the stochastic null difference, this indicates that  partner diversity is more beneficial under varying environments and therefore portfolio effect is at play.

>>>>

\section*{Abstract}
\tom{}{I think this is too long for Am Nat requirements. Also, use ``we''. I think you just pasted in an abstract that you used elsewhere, so I will work on this once you write a real abstract for the Am Nat paper.}
Mutualisms are widespread species interactions with diverse and dynamic consequences. 
They are considered more context dependent than other species interactions, meaning there are many different factors which change the outcomes of interactions between mutualists, including partner diversity. 
Partner diversity has become a central focus in the field of mutualisms, expanding previous work from primarily pairwise to multispecies mutualisms. 
It has been shown that pairwise studies are poor predictors of the effects of multispecies mutualistic interactions. 
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.

@@

\noindent Functions \tom{$P(x';\pmb{\tau^P})$ and $F(x';\pmb{\tau^F})$}{These should be $x$ not $x'$ because we modeled flowering and fertility in year t based on size in year t.} give the probability of flowering and the number of flowerbuds produced, respectively, by plants of size $x'$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit (\tom{$\kappa(a')$}{Why is this $a'$? I think it should be $a$. More generally, I think you need to explain your use of primes and how readers should interpret them at the start of model exposition.}) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that \tom{deviates are specific to plants in ant state $a$}{Why does the tau vector for viability have the $a$ subscript but the others do not?}. 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using Stan and Rstan (\tom{}{Cite Stan and RStan.}). 
Unless otherwise mentioned, all used vague priors.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given \tom{the fixed effects of previous size $x$ and previous ant partner $a$ and random effects of plot $w$ and year $u$. }{You need to go a little farther than this, because the fixed effects are arranged as a second-order polynomial with size-ant interactions. That needs to be explained and the choice of polynomial should be justified. You also do not describe the size-dependence in variance and kurtosis. Also, I would subscript the $u$ as $u_{a}$, then explain what that means. The $u_{a}$ is important to describe in a biologically meaningful way because this is how the portfolio effect is tested.  Finally, notice that the tau vector does not actually appear in the model. This is a problem. You can solve it by changing u to tau, or vice versa.}
We fit this model to size data at the end of the transition year $y^G$ using the location-scale parameterization of the \tom{student $t$ distribution}{Don't kill me but I think we should either figure out a skewed t or go back to the skew normal. But just to comment on the text as is, you need to describe the three parameters of the t (e.g., G-hat is the mean) in order for the equations to make sense.}, because in preliminary analyses we found that size transition data were more fat-tailed than a Gaussian distribution could accommodate. 
Specifically, the model was: 
$$y^G \sim Student T(\hat{nu},\hat{G},\hat{\sigma}) $$
$$\hat{G} = \beta_{0}^{G} + \beta_{1}^{G} \times x  + \beta_{2}^{G} \times a + \beta_{3}^{G} \times x \times a  + \beta_{4}^{G} \times x^2 + \beta_{5}^{G} \times a \times x^2 + u + w $$
$$\hat{\sigma} = \beta_{0}^{\sigma} + \beta_{1}^{\sigma} \times x $$
$$\hat{\nu} = \beta_{0}^{\nu} + \beta_{1}^{\nu} \times x $$
\tom{}{I think the equations will be more readable if you do not use the $\times$ symbol for multiplication.}
where \tom{$u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ with ant $a$ in year $t$}{This description is not correct. The variance is not ant-specific, the deviates are.}, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
\tom{Ants are included as predictors here because ant partners defend cholla from herbivores, preventing vegetative tissue loss.}{I think this is unnecessary since you provide the rationale for ant effects in the beginning of this section.}

\paragraph{Survival}
The survival model ($S(a,x;\pmb{\tau_{a}^{S}})$) estimates the \tom{probability of survival $y^S$}{$y^S$ is not the probability of survival} from year $t$ to year $t+1$, with fixed effects of the previous size of the cholla $x$ and ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^S \sim Bern(\hat{S})$$
$$logit(\hat{S} = \beta_{0}^{S}) + \beta_{1}^{S} \times x + \beta_{2}^{S} \times a + \beta_{3}^{S} \times x \times a + u + w$$
\tom{where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.}{Much of this is redundant with the growth model, and likely redundant with models below. Think about efficient ways to economize text so that you are not re-explaining the same things across sub-models.}

\paragraph{Reproduction}
The reproduction model ($P(x';\pmb{\tau^{P}})$) estimates the \tom{probability of reproducing $y^P$}{Same comment: $y^P$ is not the probability of reproducing.} in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^P \sim Bern(\hat{P})$$
$$logit(\hat{P} = \beta_{0}^{P} + \beta_{1}^{P} \times x' + u + w)$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Flowers}
The total flowers model ($F(x';\pmb{\tau^{F}})$) estimates the total flowers produced by a plant $y^F$ in year $t+1$, with fixed effects of size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a zero-truncated negative binomial distribution:
$$y^{F} \sim 0 Truncated Negative Binom(\hat{F},\hat{\phi})$$
$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u + w$$
$$log(\hat{\phi}) = \beta_{0}^{\phi}$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Flowerbud viability}
The viability model ($V(a;\pmb{\tau^{V}_{a}})$) estimates the \tom{proportion of flowers produced by a plant which are viable (not aborted) $y^V$}{$y^V$ is not the proportion of viable buds.} in year $t+1$, with fixed effects of ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Binomial distribution:
$$y^{V} \sim Binom(\hat{V})$$
$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau^{\epsilon}})$) estimates the probability of a cactus being visited by an ant partner $a'$ \tom{$y^{\epsilon}$}{This is not the probability of being visited.}, with fixed effects of the previous size \tom{of the cholla}{I will not make this comment everywhere but notice that ``of the cholla'' does basically no work in this sentence. It is very clear to readers at this point that ``size'' refers to size of the cholla, so these words can be cut. More generally, edit your own writing to remove words (or sentences) like this that take up space but provide no new information.} $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$, using a multinomial distribution: 
$$y^{\epsilon} ~ Multinomial(\hat{\epsilon})$$
$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$.

\paragraph{Recruit Size Distribution}
The recruit size model \tom{($n(x,a')$)}{Not the correct notation.} estimates the size distribution of all recruits \tom{$y^{\eta}$}{Think deeply about what the $y$'s represent and how you can better describe them.} from a given year $t+1$, with no fixed or random effects, using a Normal distribution: 
$$y^{\eta} ~\sim N(\hat{\eta},\hat{\sigma})$$
$$\hat{\eta} = \beta_{0}^{\eta}$$
where $\hat{\sigma}$ is estimated with a non-informative prior.

\paragraph{Germination}
With germination data from \tom{Miller et al., 2007}{CITE}, we fit two \tom{Bayesian generalized linear models}{they are all Bayesian glms so no need to say this here.} for the probability of germinating from \tom{a seed in the first year ($\gamma_1$) or the second year ($\gamma_2$)}{This can be described more accurately by referring to the 1 and 2 yo seed banks.} in year $t+1$, with no fixed or random effects, using a binomial distribution:
$$y^{\gamma_1} \sim Binomial(\hat{\gamma_1})$$
$$y^{\gamma_2} \sim Binomial(\hat{\gamma_2})$$
$$logit(\hat{\gamma_1}) = \beta_{0}^{\gamma_1}$$
$$logit(\hat{\gamma_2}) = \beta_{0}^{\gamma_2}$$

\paragraph{Pre-Census Survival}
With data collected in a \tom{2005-2006}{FIX}********** recruit census, we fit a Bayesian generalized linear model for the probability of a seedling surviving to May ($\delta$) of year $t+1$ (\tom{accounting for missed mortality events}{This could be explained more clearly, including the natural history that recruitment occurs months earlier than the census, so we need to account for mortality during that time.}), with fixed effects of the previous size $x$ and random effects of the transect $m$, using a Bernoulli distribution: 
$$y^{\delta} ~ Bern(\hat{\delta})$$
$$logit(\hat{\delta}) = \beta_{0}^{\delta} + m$$
 where $m \sim N(0, \sigma_{transect}^2)$ is the random effect of transect where the recruited individual was analyzed for survival.

\paragraph{Seeds Per Flower}
With data from \tom{Miller 2007}{CITE}********, we fit a Bayesian generalized linear model for the number of \tom{seeds produced by every flower}{This is not seeds per flower, it is seeds per fruit, since the viability model accounts for the probability that a flower becomes a fruit. I would actually move this earlier to follow that order of events.} on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$, using a \tom{Negative Binomial}{Not sure why this is written as zero-truncated.} distribution:
$$y^{\kappa} \sim 0 Truncated Negative Binomial(\hat{\kappa},\hat{\phi})$$
$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
Ant partners are included as predictors here because they reduce floral abortion rates and therefore may lead to higher numbers of seeds. 
\tom{}{This paragraph needs to address how we assigned seed number to the ``other'' category.}

\paragraph{Parameter estimation}
To obtain posterior estimates of the demographic parameters, we fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
For each model, we obtained \tom{3}{Spell out numbers less than ten.} chains of 10,000 iterations, each with randomly chosen initial conditions. 
The first 1,500 iterations were discarded. 
We did not thin the chains, thus all samples were retained following burn-in. 
We assessed parameter convergence between and within chains (\tom{Fig.}{reference figure}). 
To assess the overall model fit we carried out posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (\tom{Fig.}{reference figure}). 
\tom{All estimated parameters are available in the data. }{What does this mean?}
\tom{Data and code for all vital rate models is included in the supplemental information.}{I do not know what this refers to. Am Nat (and most journals) requires a data accessibility statement that will need to include details of data and code availability.}

\subsection*{IPM Analysis}
\tom{}{I think this section needs to start with an overview paragraph descibing in plain English what you are trying to do with the fitted model, before getting into any technical details.}
Analyzing an IPM requires discretizing the continuous IPM kernel into an approximating matrix \tom{to calculate the dominant eigenvalue}{The dominant eigenvalue is but one of many, MANY forms of ``analysis'' of an IPM. I would cut this bit that I highlighted.}. 
Size variable $x$ is discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between ant partners, leading to a matrix size of \tom{$m b \times m b$ where $m$ is the number of unique combinations of $a$ and $a'$ (how many possible ant transitions there are).}{I think what you call $m$ here is actually just $A$, the total number of ant states.} 
Our model also has two additional discrete states (year one and year two seed banks) leading to a final matrix size of $m(b+2) \times m(b+2)$.
\tom{We used $b = 200$ bins.}{It is also useful to briefly acknowledge \emph{why} 200.}
\tom{We extend the integration limits $L$ and $U$ to avoid unintentional eviction \cite{Williams2012}.}{You have not said how you estimate $L$ and $U$ from the data.}

Traditionally, in a deterministic IPM the asympotic population growth rate $\lambda$ is estimated as the dominant eigenvalue of the discrete kernel.
\tom{In a deterministic bayesian IPM,}{I thought you were only using the two types of stochastic models (and I have not read ahead).} we create \tom{1,000 discrete kernels with a unique set of parameters}{This should explicitly refer to samples from the posterior distribution of parameter values} from our bayesian statistical models to estimate 1,000 $\lambda$ values as a distribution. 
\tom{In a deterministic version of our IPM}{At this point, it is not clear to readers how many ``versions'' you have and why, which is why an overview paragraph providing a ``roadmap'' will be an important addition.}, \tom{this process is repeated separately for every combination of ant partners: complete vacancy; \textit{C. opuntiae} and vacancy; other and vacancy; \textit{L. apiculatum}, \textit{C. opuntiae}, and vacancy, \textit{L. apiculatum}, other, and vacancy; \textit{C. opuntiae}, other, and vacancy; and all ant partners and vacancy.}{Rationale for why single ant scenarios are coupled with vacancy should be better developed.}
In order to calculate stochastic $\lambda$ distributions for each ant scenario, \tom{every one of the 1,000 parameter-iteration associated $\lambda$s comes from the mean of 5,000 year-specific estimations each associated with a year random effech which was randomly selected from the 18 years of data. }{Even having worked on this with you for a long time, I cannot follow what you are saying here. This will need careful editing so that the methods and rationale for those methods (just as important!) are crystal-clear.}

\tom{We compare the distributions of $\lambda$ across each combination of ant partners to whether sampling effect or complementarity is at play in the system.}{I do not think you have sufficiently developed the logic for how these comparisons inform hypotheses about different mechanisms. Also, and this is a matter of taste, I find the language ``at play in the system'' to be not very scientific and would prefer to phrase this more precisely.}
To compare the distributions of two hypothetical \tom{populations, A and B}{It is not clear what this toy example stands in for in the real model.}, we subtract the vector of 1,000 $\lambda_A$ estimations of one population from the 1,000 $\lambda_B$ estimations of the other ($\lambda_A - \lambda_B$).
\tom{If the average of these differences is positive, population A has a higher $\lambda$ than B, if the average is 0, they are equal, and if the average is negative, population B has a higher $\lambda$ than A.}{I think this could be edited to more effectively communicate what is gained from these contrasts. The average will never be exactly zero. The important element, IMO, is that we gain a probabilistic estimate of which combinations have higher or lower fitness, given the data.}

\tom{To determine if sampling effect is at play in a system}{Ditto previous comment re: ``at play''. But it is also important to recognize, as I have emphasized before, that these are mechanisms that could explain positive diversity effects. Your writing here presumes a positive diversity effect.}, we must first determine if there is a "best" partner, by determining which single ant association is correlated with the highest \tom{$\lambda$}{Here and elsewhere, it is important to use correct notation for fitness. $\lambda$ is population mean fitness in a constant environment. $\lambda_{S}$ is population mean fitness in a stochastic environment. If you are talking about stochastic environments, you should be using $\lambda_{S}$. I also think it would be helpful to notate either $\lambda$ with ant partner, e.g. $\lambda^{C}$ is fitness corresponding to exclusive association with \textit{Crematogsaster}. I think that would help communicate methods.} estimation.
\tom{We can do this by comparing the distributions of each $\lambda$ and finding the one which is larger than all others.}{Distributions are not larger than one another. They have properties -- moments -- that can differ, like the mean or standard deviation. Be more precise here.}
Then we \tom{must show}{I don't like ``must show'' -- sounds like we are hunting for something.} that the $\lambda$ of a population with all possible partners is equal to that of $\lambda$ of a population with only the "best" partner, by comparing the distributions as described earlier.
If it is significantly larger then complementarity is potentially at play (more on this in the next paragraph).
If it is significantly smaller, this indicates that rather than positive effects of partner diversity, there are actually important costs of partner diversity that dampen the population growth rate.

@@

To determine if portfolio effect is at play we must show that partner diversity buffers the population from environmental variation.
\tom{To do this we parameterized a new set of bayesian statistical models in which year random effects were not ant specific, meaning the effects of ant partners were not able to covary with year effects (the scenario in which all ant partners must respond to annual variation the same way and portfolio effect is not possible).
The new estimations of year random effects are $u \sim N(0,\sigma_{yr}^2)$ with year specific variance of $\sigma_{yr}$.}{This needs to be explained earlier, in the statistical estimation sections, so that this section can focus on teasing out mechanisms from the model. I also think you need to better explain what the random year effects represent, biologically. We do not have any data on how the ants are actually fluctuating, but we do know something about how the effects of the ants on plants fluctuates. That is a critical nuance and it does not currently come through. I am also wondering whether it would be preferable to have the $a$ coefficients temporally fluctuate rather than making the $u$ ant-specific. Or are these effectively the same thing? (we could probably figure it out on a whiteboard)} 
With this verision of the IPM (now referred to as the stochastic null IPM) \tom{we followed the same approach to calculate distributions of stochastic $\lambda$s for each ant scenario.}{It is still not clear if what you described above was based on deterministic or stochastic $\lambda$.}
We can estimate the effects of partner diversity on the focal mutualist by calculating the difference in $\lambda$ distributions between populations with no partners and populations with all partners present.
Then we can compare the difference in the effects of partner diversity on the focal mutualist calculated from the stochastic and stochastic null IPMs respectively.
If the stochastic difference is greater than the stochastic null difference, this indicates that  partner diversity is more beneficial under varying environments and therefore portfolio effect is at play.
